Estimating Animal Abundance with N-Mixture Models Using the \pkg{R-INLA} Package for R

### 2017-04-21 meehan_et_al_inla_nmix_jss/article.tex
More fine tuning
Fixed figures

%% for pretty printing
%% comma-separated
\Plainauthor{Timothy D. Meehan, Nicole L. Michel, H\"{a}vard Rue}
%% without formatting
\Plaintitle{Estimating Animal Abundance with N-Mixture Models Using the R-INLA Package for R}
%% a short title
\Shorttitle{\pkg{R-INLA} for N-Mixture Models}

@@

}
\Keywords{wildlife, abundance, detection, N-mixture models, \pkg{R-INLA}, \proglang{R}, \proglang{BUGS}}
%% without formatting
\Plainkeywords{wildlife, abundance, detection, N-mixture models, R-INLA, R, BUGS}

@@

\begin{Code}
R> data4sim <- function(n.sites = 72,   # number study sites
+                       n.surveys = 3,  # short term replicates
+                       n.years = 9,    # number years
+                       lam.b0 = 2.0,   # intercept for log lambda
+                       lam.b1 = 2.0,   # slope for log lambda cov 1
+                       lam.b2 = -3.0,  # slope for log lambda cov 2
+                       lam.b4 = 1.0,   # slope for log lambda year
+                       p.b0 = 1.0,     # intercept for logit p
+                       p.b1 = -2.0,    # slope for logit p cov 1
+                       p.b3 = 1.0,     # slope for logit p cov 3
+                       disp.size = 3.0 # size of overdispersion
+                       ){
+  # setup
+  if(n.years %% 2 == 0) {n.years <- n.years + 1}     # make years odd
+  N.tr <- array(dim = c(n.sites, n.years))           # for abund
+  y1 <- array(dim = c(n.sites, n.surveys, n.years))  # for ex1 counts
+  y2 <- array(dim = c(n.sites, n.surveys, n.years))  # for ex2 counts
+  # abundance covariate values
+  x.lam.1 <- array(as.numeric(scale(runif(n = n.sites, -0.5, 0.5), 
+                   scale = F)), dim = c(n.sites, n.years))
+  x.lam.2 <- array(as.numeric(scale(runif(n = n.sites, -0.5, 0.5), 
+                   scale = F)), dim = c(n.sites, n.years))
+  yrs <- 1:n.years
+  yrs <- (yrs - mean(yrs)) / (max(yrs - mean(yrs))) / 2
+  yr <- array(rep(yrs, each = n.sites), dim = c(n.sites, n.years))
+  # fill abundance array
+  lam.tr <- exp(lam.b0 + lam.b1*x.lam.1 + lam.b2*x.lam.2 + lam.b4*yr)
+  for(i in 1:n.sites){
+    for(k in 1:n.years){
+    N.tr[i, k] <- rnbinom(n = 1, mu = lam.tr[i, k], size = disp.size)
+  }}
+  # detection covariate values
+  x.p.1 <- array(x.lam.1[,1], dim = c(n.sites, n.surveys, n.years))
+  x.p.3 <- array(as.numeric(scale(runif(n = n.sites*n.surveys*n.years, 
+                                        -0.5, 0.5), scale=F)), 
+                 dim=c(n.sites, n.surveys, n.years))
+  # average x.p.3 per site-year
+  x.p.3.mean <- apply(x.p.3, c(1,3), mean, na.rm=F)
+  out1 <- c()
+  for(k in 1:n.years){
+   chunk1 <- x.p.3.mean[,k]
+   chunk2 <- rep(chunk1, n.surveys)
+   out1 <- c(out1, chunk2)
+  }
+  x.p.3.arr <- array(out1, dim=c(n.sites, n.surveys, n.years))
+  # fill count array with site-yr x.p.3
+  p.tr1 <- plogis(p.b0 + p.b1*x.p.1 + p.b3*x.p.3.arr)
+  for (i in 1:n.sites){
+    for (k in 1:n.years){
+      for (j in 1:n.surveys){
+        y1[i,j,k] <- rbinom(1, size = N.tr[i,k], prob = p.tr1[i,j,k])
+  }}}
+  # fill count array with site-surv-yr x.p.3
+  p.tr2 <- plogis(p.b0 + p.b1 * x.p.1 + p.b3 * x.p.3)
+  for (i in 1:n.sites){
+    for (k in 1:n.years){
+      for (j in 1:n.surveys){
+        y2[i,j,k] <- rbinom(1, size = N.tr[i,k], prob = p.tr2[i,j,k])
+  }}}
+  # return data
+  return(list(n.sites = n.sites, n.surveys = n.surveys, n.years = n.years,
+              x.p.1 = x.p.1[,1,1], x.p.3 = x.p.3, x.p.3.mean = x.p.3.mean,
+              x.p.3.arr = x.p.3.arr, x.lam.1 = x.lam.1[,1],
+              x.lam.2 = x.lam.2[,1], yr = yr[1,], y1 = y1, y2 = y2,
+              lam.tr = lam.tr, N.tr = N.tr))
+  } #end function
R> set.seed(12345) # make results reproducible
R> sim.data <- data4sim() # create datasets
\end{Code}

@@

%% figure 1
\begin{figure}
  \includegraphics[width=\linewidth]{fig1.jpeg}
  \caption{Marginal posteriors from \proglang{JAGS} (solid gray lines) and \pkg{R-INLA} (dashed black lines), along with Maximum Liklihood estimates (black circles) and 95\% confidence intervals (horizontal black lines) from \pkg{unmarked}.  True input values are represented by vertical black lines.}
  \label{fig:fig1}
\end{figure}
%% end figure 1

@@

%% figure 2
\begin{figure}
  \includegraphics[width=\linewidth]{fig2.jpeg}
  \caption{Differences between the estimated posterior mean values and the true input parameter values as a function of the slope coefficient (effect size) for the site-survey-year covariate, x.p.3.  Black circles and lines are output from the model with the site-survey-year covariate, x.p.3, and gray circles and lines are from the model with an averaged site-year covariate, x.p.3.mean.  Parameter name and true input value is given in the strip across the top of each panel.}
  \label{fig:fig2}
\end{figure}
%% end figure 2

@@

%% figure 3
\begin{figure}
  \includegraphics[width=\linewidth]{fig3.jpeg}
  \caption{Estimates and 95\% confidence intervals from \pkg{unmarked} (gray circles and lines) and posterior means and 95\% credible intervals from \pkg{R-INLA} (black lines) from an N-mixture model of mallard duck abundance.  The \pkg{unmarked} model included site-survey covariates for survey effort and survey date, while the \pkg{R-INLA} model included site-averaged versions.  A value of zero is depicted by the vertical gray line.}
  \label{fig:fig3}
\end{figure}
%% end figure 3

@@

The key observation is that both the Poisson and the binomial distribution can be evaluated recursively in $n$,
\begin{displaymath}
    \text{Poisson}(n; \lambda) = \text{Poisson}(n-1; \lambda) \frac{\lambda}{n}
\end{displaymath}
and
\begin{displaymath}
    \text{binomial}(y; n, p) = \text{binomial}(y; n-1, p) \frac{n}{n-y}(1-p),
\end{displaymath}
and then also for the Poisson-binomial product
\begin{displaymath}
    \text{Poisson}(n ; \lambda) \; \text{binomial}(y;  n, p)
    =
    \text{Poisson}(n-1; \lambda) \; \text{binomial}(y; n-1, p)
    \frac{\lambda}{n-y}(1-p).
\end{displaymath}
If we define $f_i = \lambda(1-p)/i$ for $i=1, 2, \ldots$, we can make use of this recursive form to express the likelihood with a finite upper limit as
\begin{eqnarray}
    \text{Prob}(y) &=& \sum_{n = y}^{n_{\text{max}}}
                       \text{Poisson}(n ; \lambda)\;
                      \text{binomial}(y;  n, p) \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{binomial}(y; y, p)
                       \Big\{ 1 + f_1 + f_1f_2 +
                    \ldots
                      +f_1\cdots f_{n_\text{max}}
                       \Big\} \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{binomial}(y; y, p)
                       \Big\{ 1 + f_1(1+f_2(1+f_3(1+ \dots)))\Big\}\nonumber
\end{eqnarray}
The log-liklihood can then evaluated using the following simple \proglang{R}-code
\begin{Code}
R> fac <- 1; ff <- lambda * (1-p)
R> for(i in (n.max - y):1){
+    fac <- 1 + fac * ff / i
+  } 
R> log.L <- dpois(y, lambda, log = TRUE) + dbinom(y, y, p, log = TRUE) + 
+           log(fac)
\end{Code}
Since this evaluation is recursive in decreasing $n$, we have to chose the upper limit $n_\text{max}$ up-front, for example as an integer larger than $y$ so that $\frac{\lambda (1-p)}{n_\text{max}-y}$ is small. Note that we are computing \texttt{fac} starting with the smallest contributions, which are more numerically stable.

>>>>

%% comma-separated
\Plainauthor{Timothy D. Meehan, Nicole L. Michel, H\"{a}vard Rue}
%% without formatting
\Plaintitle{Estimating Animal Abundance with N-Mixture Models Using the R-INLA Package for R}
%% a short title
\Shorttitle{\pkg{R-INLA} for N-Mixture Models}

@@

}
\Keywords{wildlife, abundance, detection, N-mixture models, \pkg{R-INLA}, \pkg{unmarked}, \proglang{R}, \proglang{JAGS}}
%% without formatting
\Plainkeywords{wildlife, abundance, detection, N-mixture models, R-INLA, R, BUGS}

@@

\begin{Code}
R> data4sim <- function(n.sites = 72,   # number study sites
+                       n.surveys = 3,  # number replicate surveys
+                       n.years = 9,    # number years
+                       lam.b0 = 2.0,   # intercept for log lambda
+                       lam.b1 = 2.0,   # slope for log lambda cov 1
+                       lam.b2 = -3.0,  # slope for log lambda cov 2
+                       lam.b4 = 1.0,   # slope for log lambda year
+                       p.b0 = 1.0,     # intercept for logit p
+                       p.b1 = -2.0,    # slope for logit p cov 1
+                       p.b3 = 1.0,     # slope for logit p cov 3
+                       disp.size = 3.0 # size of overdispersion
+                       ){
+  # setup
+  if(n.years %% 2 == 0) {n.years <- n.years + 1}     # make years odd
+  N.tr <- array(dim = c(n.sites, n.years))           # for abund
+  y1 <- array(dim = c(n.sites, n.surveys, n.years))  # for ex1 counts
+  y2 <- array(dim = c(n.sites, n.surveys, n.years))  # for ex2 counts
+  # abundance covariate values
+  x.lam.1 <- array(as.numeric(scale(runif(n = n.sites, -0.5, 0.5), 
+                   scale = F)), dim = c(n.sites, n.years))
+  x.lam.2 <- array(as.numeric(scale(runif(n = n.sites, -0.5, 0.5), 
+                   scale = F)), dim = c(n.sites, n.years))
+  yrs <- 1:n.years
+  yrs <- (yrs - mean(yrs)) / (max(yrs - mean(yrs))) / 2
+  yr <- array(rep(yrs, each = n.sites), dim = c(n.sites, n.years))
+  # fill abundance array
+  lam.tr <- exp(lam.b0 + lam.b1*x.lam.1 + lam.b2*x.lam.2 + lam.b4*yr)
+  for(i in 1:n.sites){
+    for(k in 1:n.years){
+    N.tr[i, k] <- rnbinom(n = 1, mu = lam.tr[i, k], size = disp.size)
+  }}
+  # detection covariate values
+  x.p.1 <- array(x.lam.1[,1], dim = c(n.sites, n.surveys, n.years))
+  x.p.3 <- array(as.numeric(scale(runif(n = n.sites*n.surveys*n.years, 
+                                        -0.5, 0.5), scale=F)), 
+                 dim=c(n.sites, n.surveys, n.years))
+  # average x.p.3 per site-year
+  x.p.3.mean <- apply(x.p.3, c(1,3), mean, na.rm=F)
+  out1 <- c()
+  for(k in 1:n.years){
+   chunk1 <- x.p.3.mean[,k]
+   chunk2 <- rep(chunk1, n.surveys)
+   out1 <- c(out1, chunk2)
+  }
+  x.p.3.arr <- array(out1, dim=c(n.sites, n.surveys, n.years))
+  # fill count array with site-yr x.p.3
+  p.tr1 <- plogis(p.b0 + p.b1*x.p.1 + p.b3*x.p.3.arr)
+  for (i in 1:n.sites){
+    for (k in 1:n.years){
+      for (j in 1:n.surveys){
+        y1[i,j,k] <- rbinom(1, size = N.tr[i,k], prob = p.tr1[i,j,k])
+  }}}
+  # fill count array with site-surv-yr x.p.3
+  p.tr2 <- plogis(p.b0 + p.b1 * x.p.1 + p.b3 * x.p.3)
+  for (i in 1:n.sites){
+    for (k in 1:n.years){
+      for (j in 1:n.surveys){
+        y2[i,j,k] <- rbinom(1, size = N.tr[i,k], prob = p.tr2[i,j,k])
+  }}}
+  # return data
+  return(list(n.sites = n.sites, n.surveys = n.surveys, n.years = n.years,
+              x.p.1 = x.p.1[,1,1], x.p.3 = x.p.3, x.p.3.mean = x.p.3.mean,
+              x.p.3.arr = x.p.3.arr, x.lam.1 = x.lam.1[,1],
+              x.lam.2 = x.lam.2[,1], yr = yr[1,], y1 = y1, y2 = y2,
+              lam.tr = lam.tr, N.tr = N.tr))
+  } #end function
R> set.seed(12345) # make results reproducible
R> sim.data <- data4sim() # create datasets
\end{Code}

@@

%% figure 1
\begin{figure}
  \includegraphics[width=\linewidth]{fig1.png}
  \caption{Marginal posteriors from \proglang{JAGS} (solid gray lines) and \pkg{R-INLA} (dashed black lines), along with Maximum Liklihood estimates (black circles) and 95\% confidence intervals (horizontal black lines) from \pkg{unmarked}.  True input values are represented by vertical black lines.}
  \label{fig:fig1}
\end{figure}
%% end figure 1

@@

%% figure 2
\begin{figure}
  \includegraphics[width=\linewidth]{fig2.png}
  \caption{Differences between the estimated posterior mean values and the true input parameter values as a function of the slope coefficient (effect size) for the site-survey-year covariate, x.p.3.  Black circles and lines are output from the model with the site-survey-year covariate, x.p.3, and gray circles and lines are from the model with an averaged site-year covariate, x.p.3.mean.  Parameter name and true input value is given in the strip across the top of each panel.}
  \label{fig:fig2}
\end{figure}
%% end figure 2

@@

%% figure 3
\begin{figure}
  \includegraphics[width=\linewidth]{fig3.png}
  \caption{Estimates and 95\% confidence intervals from \pkg{unmarked} (gray circles and lines) and posterior means and 95\% credible intervals from \pkg{R-INLA} (black lines) from an N-mixture model of mallard duck abundance.  The \pkg{unmarked} model included site-survey covariates for survey effort and survey date, while the \pkg{R-INLA} model included site-averaged versions.  A value of zero is depicted by the vertical gray line.}
  \label{fig:fig3}
\end{figure}
%% end figure 3

@@

The key observation is that both the Poisson and the binomial distribution can be evaluated recursively in $n$,
\begin{displaymath}
    \text{Poisson}(n; \lambda) = \text{Poisson}(n-1; \lambda) \frac{\lambda}{n}
\end{displaymath}
and
\begin{displaymath}
    \text{binomial}(y; n, p) = \text{binomial}(y; n-1, p) \frac{n}{n-y}(1-p),
\end{displaymath}
and then also for the Poisson-binomial product
\begin{displaymath}
    \text{Poisson}(n ; \lambda) \; \text{binomial}(y;  n, p)
    =
    \text{Poisson}(n-1; \lambda) \; \text{binomial}(y; n-1, p)
    \frac{\lambda}{n-y}(1-p).
\end{displaymath}
If we define $f_i = \lambda(1-p)/i$ for $i=1, 2, \ldots$, we can make use of this recursive form to express the likelihood with a finite upper limit as
\begin{eqnarray}
    \text{Prob}(y) &=& \sum_{n = y}^{n_{\text{max}}}
                       \text{Poisson}(n ; \lambda)\;
                      \text{binomial}(y;  n, p) \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{binomial}(y; y, p)
                       \Big\{ 1 + f_1 + f_1f_2 +
                    \ldots
                      +f_1\cdots f_{n_\text{max}}
                       \Big\} \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{binomial}(y; y, p)
                       \Big\{ 1 + f_1(1+f_2(1+f_3(1+ \dots)))\Big\}\nonumber
\end{eqnarray}
The log-liklihood can then be evaluated using the following simple \proglang{R}-code.
\begin{Code}
R> fac <- 1; ff <- lambda * (1-p)
R> for(i in (n.max - y):1){
+    fac <- 1 + fac * ff / i
+  } 
R> log.L <- dpois(y, lambda, log = TRUE) + dbinom(y, y, p, log = TRUE) + 
+           log(fac)
\end{Code}
Since this evaluation is recursive in decreasing $n$, we have to choose the upper limit $n_\text{max}$ up-front, for example as an integer larger than $y$ so that $\frac{\lambda (1-p)}{n_\text{max}-y}$ is small. Note that we are computing \texttt{fac} starting with the smallest contributions, which are more numerically stable.
